---
title: "Reproducible Research: Peer Assessment 1"
output:
  md_document:
      variant: markdown_github
html_document: PA1_template.html
keep_md: true
---


## Loading and preprocessing the data
The activity data is downloaded from the assessment link provided in the REAdME.md file.  
In addition the packages that are used are loaded into the environment

```{r activity data loading, results='hide', warning=FALSE}
#checks if activity object exists, if not downloads the data and extracts from the zip file 
if (!exists('activity')) {
  destfile <- "activity.csv"
  url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(url, "repdata-data-activity.zip", mode="wb")
  unzip("repdata-data-activity.zip")
  activity <- read.csv(destfile)
  remove(destfile)
}

#library package loading
p <- c('plyr', 'ggplot2', 'Rcpp')
lapply(p, library, character.only=T)
```


## What is mean total number of steps taken per day?

This assessment requires the calculation of the total number of steps per day. i.e. sum the data across the day and take the mean across all days  
Additionally a histogram for the intraday steps (across the intervals each day) is produced.
```{r daily steps analysis}
#summary analysis of the data to calculate the total, mean, and median daily steps
dailysteps <- ddply(activity, .(date) , summarize, daysteps =sum(steps))
total_steps <- sum(activity$steps, na.rm=TRUE)
mean_daily_steps <- round(mean(dailysteps$daysteps, na.rm = TRUE),2)
median_daily_steps <- median(dailysteps$daysteps, na.rm = TRUE)

#create a histogram of the mean daily steps taken
qplot(daysteps, data=dailysteps, geom='histogram')
png("figure/dailyanalysis.png")
qplot(daysteps, data=dailysteps, geom='histogram')
dev.off()

#print out the summary data for the analysis
summary <- data.frame(total_steps, mean_daily_steps, median_daily_steps)
print(summary)

```


## What is the average daily activity pattern?

Creates a single line graph displaying the steps take across the day's interval pattern
From this and assessment of the maximum steps taking in a individual interval is take and printed to the display
```{r daily pattern (intraday)}
#timeseries plot of 5min interval steps taken
intradaysteps <- ddply(activity, .(interval) , summarize, steps =mean(steps, na.rm = TRUE))
qplot(interval, steps, data=intradaysteps, geom="line")
png("figure/intraday.png")
qplot(interval, steps, data=intradaysteps, geom="line")
dev.off()

#return the 5min interval with the maximum average steps each day
sortedinterval <- intradaysteps[rev(order(intradaysteps$steps)),]
sortedinterval[1,]
```


## Imputing missing values
The missing values are replaced with the mean for the time interval across the data set. This result is then used for the subsequent analysis.
A repeat of the daily analysis in Question 1 is repeated on the NA cleaned data.
```{r NA removal}
countNA <- sum(is.na(activity$steps))

#replace the NA fields with the mean for that time interval using the optmatch package
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
activityNA <- ddply(activity, ~ interval, transform, stepsna = impute.mean(steps))

#copy of steps from question 1 replacing the data set for the updated NA cleaned data
dailystepsNA <- ddply(activityNA, .(date) , summarize, daysteps =sum(stepsna))
mean_daily_stepsNA <- round(mean(dailystepsNA$daysteps, na.rm = TRUE),2)
median_daily_stepsNA <- median(dailystepsNA$daysteps, na.rm = TRUE)
qplot(daysteps, data=dailystepsNA, geom='histogram')
png("figure/dailyanalysisNA.png")
qplot(daysteps, data=dailystepsNA, geom='histogram')
dev.off()

#print out the summary data for the analysis
summaryNA <- data.frame(countNA, mean_daily_stepsNA, median_daily_stepsNA)
print(summaryNA)
```


## Are there differences in activity patterns between weekdays and weekends?
A new 2 factor flag ('weekendflag') is created to separate weekdays from weekends. This field is used to create a comparasion between the steps pattern for weekdays versus weekends
```{r weekday variance, warning=FALSE}
#calculated the difference between weekday and weekend activity
#3 part clean up process 
#   1. Convert the date field to a date class
#   2. Create the  factor flag for weekend/weekday
#   3. Convert the weekend flag to a factor class
activityNA$date <- as.Date(activityNA$date)
activityNA$weekendflag <- ifelse(weekdays(activityNA$date) %in% c('Saturday', 'Sunday'),'weekend', 'weekday') 
activityNA$weekendflag <- as.factor(activityNA$weekendflag)

# Create the 2 level lattice plot using ggplot2
qplot(interval, steps, facets = weekendflag~., data=activityNA, geom= 'line')
png("figure/weekdayNA.png")
qplot(interval, steps, facets = weekendflag~., data=activityNA, geom= 'line')
dev.off()
```

The comparasion line graphs show weekdays have higher step counts across the day, especially earlier in the day. While weekends have a similar period maximum, it happens later in the day and that maximum is not sustained for as long.